---
title: "Neuroimmunolgical-methods-in-stroke-research-course-microbiota-analysis"
output: html_document
date: "2023-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Neuroimmunolgical-methods-in-stroke-research-course


### Mathematical operations 

Addition
```{r}

```

Subtraction 
```{r}

```

Multiplication 
```{r}

```

Division
```{r}

```

Modulo 
```{r}

```

Square
```{r}

``` 

### Basic types 

There are 6 basic data types in R: logical, numeric, integer, character, raw and complex. Today (and in general) only the first four are important. 

```{r}
class(TRUE)
class(3.0)
class(6L)
class("Hello")
```

### Data structures 

Four most commonly used data structures 

#### 1D
Vectors - ordered collection of data of a given length and all of the same type:
```{r}

```

Lists - ordered collection of data/objects of a given length. Can be composed of heterogenous types. 
```{r}

```

#### 2D 

Matrices - 2D data structure (rows x columns) of the same type
```{r}

```

Dataframes - 2D heterogenous data structure (rows x columns)
```{r}

```

### Variable assignment 

Variables are assigned in R using the ```<-``` operator. A variable can be though of a a "box" you store a value in for later use.

They can be used to store a value which you want to be constant throughout your code e.g. a file name 
```{r}

```

Or to save time by storing long strings/numbers 
```{r}

```

Objects such as vectors and dataframes can also be stored in variables as you saw above. ```df``` and ```vector2``` are both variables. 

### Control flow

As well as writing individual statements like we have done above we can also use logic to control the flow of our code, allowing us to repeat bits of code or only run if a given condition is met. 

for loops allow us to repeat code a specified number of times e.g.

```{r}

```

This prints out each element in the ```vector2```variable we defined earlier. Not particularly interesting though.. 
We will discover a more interesting use case later. 

if/else statements allow us to control the flow of our code better: 

```{r}

```

Change the values of x and y and see how the output changes. 

### Functions 

Functions are used to abstract repetitive code in something which is reusable and modular, making code easier to read and debug. 

Using the above if else example we could create a function called ```greater_than``` which tells us which value is highest in a more modular way
```{r}

```

```{r}

```

More interesting functions might perform a calculation for us and return the value 

```{r}

```

```{r}

```

You can also use built-in functions or install packages to access functions written by others. It's usually the case that someone has already written a function for whatever you want to do. 
External packages can be downloaded using the ```install.packages``` function and loaded using the ```library``` function. 

If you need help with a function you can also type ?functionname in the console e.g. ?log10 and the help for that function will show up, detailing what the function does, what inputs it expects and what value(s) it returns. 

### Data wrangling 

A few key concepts on loading and manipulating data.
```{r}

```


Now we understand a little bit of R we can move on to the actual analysis. 

# Analysis of Microbiota data

## Data Analysis 

```{r}

```

Firstly, we will read the ASV table(s) and metadata into R, using
phyloseq. We have included the data used in this publication to
demonstrate our analysis pipeline.

```{r}

```

Phyloseq is used as it provides a nice way of storing all the associated
data in one object or class. As you can see, the metadata, tree and ASV
table are all combined here into `ps`. Importantly, each part can also
be accessed invidivually.

## Data Normalisation

Here will transform the data using minimum sum normalisation and
relative abundance for downstream analyses.

### How does normalisation work?

Minimum sum normalisation works by dividing the sum in each sample by
the minimum sum across all samples, thereby standardising each sample to
a fixed sum.

Relative, transforms each sample into compositions, on a fixed scale of
0-100.

```{r}

```

We can see how this works by looking at the column sums:

```{r}

```

```{r}

```

## Alpha Diversity

Here we will calculate three different measures of alpha-diversity:

-   Species richness, or the number of observed species
    -   This is calculated by
-   Shannon effective diversity, measuring
-   Faith’s PD, measuring the phylogenetic component of alpha diversity

The function `calc_alpha` wraps all of these calculations and only
requires the mss normalised phyloseq object as input, returning a
dataframe with a column for each above-mentioned dataframe.

```{r}

```

### How are alpha diversity metrics calculated?

Richness here is calculated as the the total number of observed species
greater than 0.5 mss normalised abundance. This threshold is used to
exclude noise (see Lagkouvardos et al 2017, PeerJ and Reitmeier et al
2021, ISME Communication for a more thorough explanation).

Shannon effective diversity is calculated as the exponent of the Shannon
index:

$$H = -\\sum\_{i=1}^{R} p\_i ln(p\_i)$$
where *R* = richness, *p<sub>i</sub>* is the relative abundance of
species *i* and *ln* is the natural logarithm.

This metric accounts for both the abundance and evenness of taxa.

Lastly, Faith’s alpha diversity is calulated as the sum of the branch
lengths in the phylogenetic tree. This method does not account for
abundance, only measuring the phylogenetic diversity of a sample.

Now we have calculated alpha diversity, we can merge this information
with our metadata to generate a dataframe for plotting.

```{r}

```

### Plotting

To plot the alpha diversity metrics we will use a boxplot with jittered
points layered on top. The function `plot_boxplot`will do this for you,
we just need to set some parameters first.

Firstly, we will list the statistical comparisons we want to make, by
creating a list of vectors. In this case we only have two groups, stroke
and sham which can be written like: `list(c("Stroke", "Sham"))`. If we
had an extra group, for example, “Control”, we would then have three
comparisons to define like so:
`list(c("Control", "Sham"), c("Control", "Stroke"), c("Sham", "Stroke"))`

Comparisons list

```{r}

```

We can also specify the colours we want to use in our plots here by
creating a named vector of colours.

```{r}

```

To generate the plot we need to provide the dataframe, the name of the
column containing the grouping variable (in this case simply “Group”),
the name of the column containing the values to plot (Richness here). To
colour by group we provide the column name of the grouping variable to
`fill_var`. We can then add the list of comparisons, some x and y-axis,
a title if desired and the plot colours.

In instances where the alphabetical order of your group levels does not
match the order you would like to plot them in, you can specify this
here with the `group.order`parameter.

#### Richness

```{r}

```


```{r}

```

#### Faith’s PD

```{r}

```


Statistical significance is calculated internally in the `plot_boxplot`
function using unpaired wilcoxon, thus currently this function is only
suitable when groups are independent. Note that if no significance is
displayed on the plot, the the differences between groups were not
statistically significant.

## Beta Diversity

Here we will calculate beta-diversity based on Generalized unifrac
distance and plot an ordination of this using Non-metric
multidimensional scaling.

The `calc_betadiv` function calculates a distance matrix, and an
ordination of that matrix, returning both as a list.

Various dissimilarity indices are available:

-   Bray-Curtis - A count-based dissimilarity metric (beta-diversity),
    based on the fraction of overabundant counts.
-   Unifrac - A phylogenetic beta-diversity metric measuring the
    fraction of unique branches in a phylogenetic tree.  
-   Weighted Unifrac - An extensions of unifrac taking abundance into
    account in addition
-   Generalized Unifrac - A further extension of unifrac, placing less
    weight on highly-abundant lineages.

Similarly, there are also various ordination options:

-   NMDS (default) - Non-Metric Multidimensional Scaling. An ordination
    method which attempts to represent the dissimilarity between
    samples, as closely as possible in a low-dimensional space.
-   MDS/PCoA - Principal Coordinate analysis (also known as Metric
    Multidimensional Scaling). An ordination method which attempts to
    preserve distance between samples in a low dimensional Euclidean
    space.

```{r}

```

To plot beta diversity a convenience function `plot-beta_div` is
provided. We just need to provide the phyloseq object, the ordination
and distance matrix from the betadiv object above (This object is a list
so we can access elements with $ notation and their name), a grouping
variable and again the colours.

Since the ordination under the hood returns MDS as the column names,
regardless of whether metric or non-metric dimensional scaling is used,
we can adjust the axis labels by adding new x and y labels to our plot.
All plotting functions return ggplot2 objects which can be added to and
further customised as desired.

Within this function statistical testing of group separation is also
carried out using the adonis function of vegan. This function performs a
Permutational Multivariate Analysis of Variance or PERMANOVA test. The
resulting R<sup>2</sup> and p-value are added to the plot in the bottom
left.

```{r}

```

\#\# Differential abundance

The final step of this pipeline is to calculate differentially abundant
taxa between conditions.

This function performs the ancom-bc test, a compositionally aware
differential abundance method and returns significant results. As input,
only the phyloseq object and the column name of the grouping variable is
required.

```{r}

```

To visualise differentially abundant taxa, we provide a function which
calculates fold change of significant taxa from above and plots
diverging dot plot coloured by group, providing a clear figure showing
which taxa change between conditions.

To this function, we need to provide the results of the ancom test
above, an ordered vector of the group levels e.g. `c("Sham", "Stroke")`.
Additionally, we can provide the group colours to make interpretation
easier.

```{r}

```

